\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles

### 2019-07-10 IPANEMAP-2019.tex

\noindent \OurTool{} combines statistical sampling, unsupervised machine learning, and multi-objective optimization, to produce secondary structure models that are both stable and well-supported by experimental evidences. An analysis of several datasets demonstrates the performance of \OurTool, even in a mono probing setting, and confirms the potential of integrating multiple sources of probing data, leading to several recommendations for the experimental design of informative probing assays.

@@

However, independent computational predictions for the same RNA using probing data obtained with different probes often yield largely different models, none of which are fully consistent with all the probing data. It follows that, while theoretically informative, a multiprobing strategy often leaves a user with different models, from which one cannot objectively decide. In order to identify the best fit with all the data, researchers resort to very intuitive and manual methods such as projecting the results obtained with one probe onto the different models obtained using the constraints from another probe~\cite{Herbreteau2005,James2008,Weill2004}. At the end of the modeling process, the modeler is often left with several alternatives, all of which may appear equally consistent with the probing data, and the choice of a "best" model becomes somewhat subjective. Here we have developed a new modeling procedure that takes into account multiple probing data and ultimately yields only one or two models.

@@

\subsection*{The \OurTool{} method}
We introduce the Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles (\OurTool{}), a novel approach that integrates the signal produced by various chemical probes. It performs clustering based on multiple sets of structures sampled in different experimental conditions, to ultimately return a set of structures that are representative of the dominant clusters. Its underlying rationale is that the prominent presence of a stable secondary structure within the  (pseudo-)Boltzmann ensembles induced by multiple experimental conditions should increase its likelihood to be (one of) the native structure(s) for a given RNA. It is thus hoped that integrating several reactivity profiles may be used to promote the native structure as one of the dominant structures within the multi-ensemble, and help circumvent the limitation of pseudo-energies derived from single reactivity profile, which are generally not sufficient to elect the native structure as its minimum (pseudo)-free energy candidate. In other words, combining ensembles of structures generated using multiple probing experiments is likely to denoise the Boltzmann (multi-)ensemble, and thus mitigate systematic biases induced by experimental conditions and reagents.

@@

Since \OurTool{} supports any number of probing profiles, we consider the Cordero~\emph{et al} dataset, further described. It consists of 6 RNAs of known structures, for which reactivity profiles are available for the CMCT, NMIA and DMS reagents. We executed \OurTool{} with default parameters on each sequence and any subset of the three available conditions. The centroid secondary structure  associated with the largest probability cluster was considered as the final prediction. Predictions were evaluated in term of the geometric mean (GM) metrics, for the sake of comparison with previous studies~\cite{Spasic2017}. As a control experiment, we also report ${\tt RNAfold}$ predictions in presence/absence of probing data, both in energy minimization (MFE) and maximum expected accuracy (MEA) modes. Figure~\ref{cordero1} shows the averaged GM values for all combinations of tools and probing data.

@@

Our analyses reveal that integrating multiple profiles of at least three conditions allows to mitigate the effect of outlier conditions and, at times, to produce better secondary structure models than the one inferred only from the most accurate profile. A comparison of the joint performances to the average appears fair since no single reagent appears to systematically dominate in term of predictive capacity. It follows that a modeler cannot objectively favor \emph{a priori} a probing condition/reagent~\cite{Yu2018}. In other words, multiple experimental probing data, even combined in a pairwise fashion, were typically observed to mitigate the empirical risk of a misprediction.

>>>>

\noindent \OurTool{} combines statistical sampling, unsupervised machine learning, and multi-objective optimization, to produce secondary structure models that are both stable and well-supported by experimental evidences. The analysis of several reactivity, both publicly available and specifically produced experimentally, demonstrates the good performances of \OurTool, even in a mono probing setting. It confirms the potential of integrating multiple sources of probing data, leading to recommendations for the experimental design of informative probing assays.

@@

However, independent computational predictions for the same RNA using probing data obtained with different probes often yield largely different models, none of which are fully consistent with all the probing data. It follows that, while theoretically informative, a multiprobing strategy often leaves a user with different models, from which one cannot objectively decide. In order to identify the best fit with all the data, researchers resort to very intuitive and manual methods such as projecting the results obtained with one probe onto the different models obtained using the constraints from another probe~\cite{Herbreteau2005,James2008,Weill2004}. At the end of the modeling process, the modeler is often left with several alternatives, all of which may appear equally consistent with the probing data, and the choice of a "best" model becomes somewhat subjective. Here we have developed a new modeling procedure that jointly takes into account multiple probing data, and ultimately yields a small collection of secondary structure models.

@@

\subsection*{The \OurTool{} method}
We introduce the Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles (\OurTool{}), a novel approach that integrates the signal produced by various chemical probes.

It performs a structural clustering across multiple sets of structures sampled in different experimental conditions, and ultimately returns a set of structures that are representative of the dominant conformations supported across conditions. Its underlying rationale is that the prominent presence of a stable secondary structure within the  (pseudo-)Boltzmann ensembles induced by multiple experimental conditions should increase its likelihood to be (one of) the native structure(s) for a given RNA. It is thus hoped that integrating several reactivity profiles may be used to promote the native structure as one of the dominant structures within the multi-ensemble, and help circumvent the limitation of pseudo-energies derived from single reactivity profile, which are generally not sufficient to elect the native structure as its minimum (pseudo)-free energy candidate. In other words, combining ensembles of structures generated using multiple probing experiments is likely to denoise the Boltzmann (multi-)ensemble, and thus mitigate systematic biases induced by experimental conditions and reagents.

@@

Since \OurTool{} supports any number of probing profiles, we consider the Cordero~\emph{et al}~\cite{Cordero2012a} dataset, further described in the previous section. It consists of 6 RNAs of known structures, for which reactivity profiles are available for the CMCT, NMIA and DMS reagents. We executed \OurTool{} with default parameters on each sequence and any subset of the three available conditions. The centroid secondary structure  associated with the largest probability cluster was considered as the final prediction. Predictions were evaluated in term of the geometric mean (GM) metrics, for the sake of comparison with previous studies~\cite{Spasic2017}. As a control experiment, we also report ${\tt RNAfold}$ predictions in presence/absence of probing data, both in energy minimization (MFE) and maximum expected accuracy (MEA) modes. Figure~\ref{cordero1} shows the averaged GM values for all combinations of tools and probing data.

@@

Our analyses reveal that integrating multiple profiles of at least three conditions allows to mitigate the effect of outlier conditions and, at times, to produce better secondary structure models than the one inferred only from the most accurate profile. A comparison of the joint performances to the average appears fair since no single reagent appears to systematically dominate in term of predictive capacity. It follows that a modeler cannot objectively favor \emph{a priori} a probing condition/reagent~\cite{Yu2018}. In other words, multiple experimental probing profiles, even combined in a pairwise fashion, were typically observed to mitigate the empirical risk of a misprediction, and should be considered.
